---
title: "R basics - Morning Session"
author: "FAES-AABI"
date: "August 8, 2018"
output:
  html_document:
    df_print: paged
    toc: yes
    number_sections: no
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,tidy=TRUE, dpi=300)
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
```


## 2.3. Programming with R
R programming consists of methods for manipulation of data stored in objects (variables). The key components of R language environment are:  

- Variables/Objects
- Operators
- Conditions
- Loops
- Functions
- Models
- Scripts

### 2.3.1 Variables/Objects  
A variable is a symbolic identifier that holds particular data (reserved memory locations).  Variables can hold different types of values, i.e., integer (1,2,3), numeric (0.1,2.5,3e-11), complex (3+2i), character ('red','blue'), and logical (TRUE, FALSE), which are the five basic data classes in R ("atomic classes"). The variables are assigned with R-objects that become the data type of a variable. Based on the data type of a variable, the operating system allocates memory and decides what can be stored in the reserved memory. Most frequently used R-objects are:  
- Vectors  
- Matrices  
- Data Frames  
- Factors  
- Lists  

## Rules for proper variable names
- can be a combination of letters, digigts, period {.} and underscore {_}
- must start with a letter or a period; if it starts with a period, it cannot be followed by a digit
- operators like {* - + < > = & | %}, punctuation like {( ) { } , ; [ ]}, the comment character {#} or reserved words like {TRUE FALSE NULL NA} cannot be used as identifiers  

Valid: var_name, .varName_1, varname, VAR  
Invalid: .1name, TRUE, _varName, 1varname, var_name%, 'var Name'

Note: When data are read into R and any column names are not valid, R will change them by prefixing {X} when the first position is not valid and replacing each invalid character with a {.}; the result can be checked with the *make.names()* function, i.e., make.names('100*X (%)')

## Variable assignment
Variables can be assigned values using either the {leftward}, {rightward}, or {equal to} operator. Values can be output by typing the variable name into the R console. Alternatively, *print()* or *cat()* can be used to combine additional components or to allow printing from within a sourced script.


```{r}
# leftward operator
var_1 <- 5

# equal to operator
var_1 = 5

# rightward operator
5 -> var_1

# output variable values
var_1
print(var_1)
cat('Variable (1) =', var_1, '\n')

# assignment and output
(var_1 = 5)

```
## A. Vectors
A vector in R is a sequence of values. Special cases of vectors are scalars (vectors of length 1) and empty vectors. Function *c()* concatenates elements of a vector.

```{r}
# Create a numeric vector and output assigned variable values
x_1 = c(2, 5, 6, 10)
print(x_1)

# Create a character vector and output asigned variable values
x_2 = c('red','blue','blue','orange')
print(x_2)

# Check class
class(x_1)
class(x_2)

# check length
length(x_1)
length(x_2)

# calculate mean of numeric vector
(mx1 = mean(x_1))

# check if there are duplicated values
duplicated(x_2)
any(duplicated(x_2))

# list unique values
unique(x_2)

# table values
table(x_2)

# summarize values
print(summary(x_1))
print(summary(x_2))

# check attributes and assign names to values of the numeric vector
attributes(x_1)
names(x_1) = c('c1','c2','c3','c4')
print(x_1)
attributes(x_1)

# slice a vector of length 1 (scalar) by index or name
(s_1 = x_1[4])
(s_1 = x_1['c4'])

# create empty vectors and check their classes
e_0 = NULL
print('e_0: values, class, length')
e_0; class(e_0); length(e_0)

e_3 = vector(mode='numeric', length=3)
print('e_3: values, class, length')
e_3; class(e_3); length(e_3)

c_0 = vector(mode='character')
print('c_0: values, class, length')
c_0; class(c_0); length(c_0)

```
## B. Matrices
A matrix in R is a collection of data values arranged in a two-dimensional table with rows and columns as the 1st and 2nd dimension, respectively. All columns in a matrix have to be of the same type (.e., if a mixture of numeric and character values are given, R assigns character mode to all the values). Matrices with numeric elements are mostly used though.

```{r}
# create a matrix variable (numeric), specifying number of rows, columns, and fill dimension)
values = 1:9
names(values) = paste('v', 1:length(values), sep='.')

m_1 = matrix( values, nrow = 3, ncol = 3, byrow = TRUE )
print(m_1)

m_2 = matrix( values, nrow = 3, ncol = 3, byrow = FALSE )
print(m_2)

# matrix length and names
length(m_1)
names(m_1)
names(m_1) = names(values)

# 2-way dimensions
nrow(m_1)
ncol(m_1)
dim(m_1)

# dimnames
rownames(m_1) = paste('row', 1:nrow(m_1), sep='.')
colnames(m_1) = paste('col', 1:ncol(m_1), sep='.')
m_1
dimnames(m_1)

# summary
summary(m_1)
str(m_1)

# calculate total mean, mean by rows or columns
mt=mean(m_1)
mt
mr = rowMeans(m_1)
print(mr)
mc = colMeans(m_1)
print(mc)

# slice matrix
m_1a = m_1[1:2, 2:3]
m_1a
m_1b = m_1[, c('col.3','col.2')]
m_1b

# create matrix from numeric and character values
mix = c(1:6, 'A','B','C')
m_2 = matrix(mix, nrow =3, ncol=3, byrow=TRUE)
m_2

# check class and mode of numeric and character matrix
class(m_1)
class(m_2)
mode(m_1)
mode(m_2)

```

## C. Data Frames
A data frame in R is a two-dimensional table-like structure with rows and columns, but with different properties than a matrix. In a data frame each column is a variable, and variables can be of different type, that is of numeric, character or factor type (explained below in 6.6). When data are read into R from external sources they are assigned data frame type by default.  

Example data frame:
Boeckmann, Sheiner and Beal (1994) report data from a study by Dr. Robert Upton of the kinetics of the anti-asthmatic drug theophylline. Twelve subjects were given oral doses of theophylline then serum concentrations were measured at 11 time points over the next 25 hours. *help(Theoph)*

```{r}
# load internal R data and assign a simpler object
data(Theoph)
df = data.frame(Theoph)

# check data type and properties
str(df)
class(df)
dim(df)
length(df)
names(df)
colnames(df)
head(df)

# slice data frame
class(df[,5])
class(df[,5, drop=FALSE])

df[1:5,'conc']
df$conc[1:5]

# subset data frame by a condition
df_1 = subset(df, df$Subject == 1)

# add new variable
#.measuremnt number
df_1$num = 1:nrow(df_1)
#. difference to baseline at time 0
df_1 = data.frame(df_1, dif=df_1$conc - df_1$conc[1])

# plot variables
plot(df_1$Time, df_1$conc, type='l', main='Patient #1')

```

## D. Factors
A factor in R is a categorical variable that stores unique values of the variable as levels (integers or strings). Factors help with grouping of data and are used in statistical modeling.

```{r}
f_1 = rep(1:4, 4)
f_1
f_1 = factor(f_1)
f_1

f_2 = rep(c('G1','G2','G3','Control'), 4)
f_2
f_2 = factor(f_2)
f_2

x = rnorm(16, mean=20, sd=2.5)
mx = tapply(x, f_2, mean)
mx

```

## E. Lists
A list in R is an object which can store data of different types and dimensions.

```{r}
# create a list
l_1 = list(data1=x_1, data2=c(F,F,T), data3=m_1, data4=df_1[1:4,])
l_1

# subset list
l_2 = l_1[1]
class(l_2)

# access variable in a list
dat = l_1[[1]]
dat = l_1$data1
class(dat)

# split data frame
list_1 = split(df, paste(df$Subject))[1:2]
list_1

```

#	2.3.2. Operators
Operators are symbols representing specific mathematical or logical operations performed on data. R can perform basic operations like addition, multiplication, division etc., along with a host of more advanced manipulations. Several operations work similarly on different types of variables.


```{r}

## Arithmetic operators

# add two numbers
20+30

# multiply two numbers stored in variables s1, s2
s1 = 20; s2 = 30
s1*s2

# divide two vectors v1,v2
v1 = sample(5); v2=sample(5)
v1; v2; v1/v2

# element wise operation
sqrt(v1); v1^3

# sum all elements of a vector
sum(v1)
cumsum(v1)

# apply an operation on row/columns of a matrix/data.frame
m_1
apply(m_1, 2, log2)
apply(m_1, 1, log2)

## Logical operators

# compare elements of two variables (greater than or equal to)
v1 >= v2

# OR/AND, not equal
v1 >= 2 | v2>=3
v1 >= 2 & v2 != 3

# is TRUE/FALSE
any(v1<2)
all(v1<2)
!isTRUE(2>0)
```

#	2.3.3 Conditions
Condition test if a statement is true or false

```{r}
#check if s1 is greater than s2
#if the condition is true, assign a value for s3 as s2-1
#if the condition is false, assign a value for s3 as s2+1

if (s1 > s2) { 
  s3 = s2-1 
  } else { 
  s3 = s2+1 
}
s3

# assign category based on a condition with ifelse() function
v3 = ifelse(v1 > v2, 'red', 'blue')
v3

```

# 2.3.4 Loops
Loops are used to do a particular operation on a set of data, without having to manually perform that actions multiple times

```{r}
# the for loop goes over each number from 1 to 10, assigning the current number in the range to i, for every iteration. A cat() statement is then executed for each of the interation
for(i in 1:10)
{
  cat(i, ', ', sep='')
}

# the same for loop but with an addition statement
for(i in 1:10)
{
  xfor=i+1
  cat(xfor, ', ', sep='')
}
```

# 2.3.5 Functions
Functions are the crucial part of R. Functions are a set of in-built commands that perform certain standard tasks. For example, 'print' is a function that prints the values of the variable. 'sum' is a function that sums up the values that are passed to it, etc. Below are some of the important functions to remember;

- *help()*
-	*example()*
-	*demo()*
-	*vignette()*
-	*args()*
-	*length()*
-	*range()*
-	*toupper()*
-	*tolower()*
-	*mean()*
-	*median()*
-	*sum()*
-	*print()*
-	*dim()*
-	*mode()*
-	*class() *
-	More... http://www.columbia.edu/~cjd11/charles_dimaggio/DIRE/resources/R/rFunctionsList.pdf  

You can also write your own functions, for chunks of codes that you repeatedly use. This way, instead of typing all of the lines of the code every time, you could just call the function and be done with that. It's like magically using a single word to run a bunch of lines of code.

```{r fig.width=12, fig.height=12}
#write a function plot.patient() to produce a pharmacokinetics plot for each patient in the Theoph data set

plot.patient <- function(input=NULL, colo='black', ...){
  
  if(is.null(input)){
    data("Theoph")
    df = Theoph
  } else {
    df = input
  }
  
  patientID = paste(df[,'Subject'])
  list_df = split(df, patientID)
  
  par(mfrow=c(4,3))
  for(i in 1:length(list_df)){
    x = list_df[[i]]$Time
    y = list_df[[i]]$conc
    plot(x,y, type='o', pch=16, col=colo, xlab='time (hrs)', ylab='blood (mg/L)')
    title(paste('PatientID =',i))
}
}

#plot.patient()
plot.patient(colo='blue')

```

2.4. Scripting
Scripting is a task of placing a bunch of commands that perform individual actions, in a single file and executing all of those commands by running that one single file in a automated way.
Below is an example script that reads the interactions files, grabs a subset of the file content, filters it for the score, plots a graph of the score and writes out a new file with the filtered set of interaction.
Place all of these commands in one single text file (using a text editor like notepad or textedit), save it as a script (use .r or no extension).

```{r}
#read the input data
stat1_interactions_new <- read.csv("~/stat1_interactions.tsv",sep="\t")

#extract the first two and the last column
statsmallnew <- stat1_interactions_new[,c(1,2,15)]

#extract the interactions with more than 0.5 score
statsmallnewfiltered <- subset(statsmallnew, combined_score >= 0.5)

#write the filtered interactions in to a file
write.table(statsmallfiltered, "statsmallfiltered.txt", sep = "\t", quote = FALSE, row.names = FALSE)

#print the score distribution graph
png('statsmallnew.png')
plot(statsmallnew$combined_score)
dev.off()

message("Script ran successfully, all outputs are in the working director")

#If the above script is named as "interaction-data-process-script.r", run the script in the console using the following command;
#source('~/interaction-data-process-script.r')
```


### 2.5 RMarkdown
*RMarkdown* allows to create documents that serve as a neat and reproducible documentation of an analysis. *RMarkdown* is a very simple 'markup' language which provides methods for creating documents with headers, images, links etc. from plain text files, while keeping the code alongside its output (graphs, tables, etc.) with conventional text to explain it, a bit like a notebook. Markdown documents can be converted to other file types like .html , .pdf, or .doc

**To create a new RMarkdown file (.Rmd), select File -> New File -> R Markdown in RStudio, then choose the file type (HTML here)**

- Table
```{r }
load("C://Users/YO/Documents/example_data.RData")
tab = dat_chol

library(DT)
datatable(tab, colnames=names(tab), options = list(scrollX = FALSE, keys = TRUE, pageLength = 10),  caption=paste("Table 1"), rownames = TRUE)

```

- Figure
```{r fig.width=2, fig.height=2, dpi=300}
load("C://Users/YO/Documents/data_shapes.rda")
dat = dat_wide2
cormat = cor(dat)
library(heatmaply)
heatmaply(cormat, cexRow = 0.5, cexCol=0.5)
```

#### 2.5.1 Code chunk instructions

#### 2.5.2 Formatting text

*Italic*

**Bold**

* Unordered list item

1. Ordered list item

[Link to RStudio RMarkdown](https://rmarkdown.rstudio.com/index.html)




